Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-03-14 output/paper/manuscript.qmd
Re-writing discussion basically.

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} was associated with attentional function, and with total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to attentional function. The literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of non-persistent {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment (too broad) in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].

>>>>

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to certain non-persistent {{< acr edcs >}} was associated with attentional function (MEPA, MEHP, oh-MiNP, and oxo-MiNP), and with total production of cortisol, cortisone, and corticosterone (DEP, DMP, DMTP, BPA, ETPA, MEPA, MEHP, oh-MiNP, and oxo-MiNP). Increased production of these glucocorticosteroids did not seem to affect attentional function. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

===
To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.

More generally, the literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @LiPapandonatosCalafat:2019; @BalalianWhyattLiu:2019; @ShoaffCoullWeuve:2020; @JankowskaPolanskaHanke:2019; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019].

González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@wechsler1949wechsler] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.

Preliminary evidence is also available for several phthalate metabolites in relation to cognitive development in childhood. Higher levels of di(2-ethylhexyl) phthalate metabolites (including MEHP, MEHHP, and MEOHP) were associated with lower intelligence scores in children aged 2 to 12 years [@HuangChenSu:2015], lower scores of {{< acr iq >}} and verbal {{< acr iq >}}, more omission errors (a measure of inattention), and higher scores of response time variability (a measure of sustained attention) in 6-year old Korean children [@KimHongShin:2017], and poorer fine motor skills in preadolescent boys [@BalalianWhyattLiu:2019].

Few studies have looked into different classes of non-persistent {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment (too broad) in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].
===
